Binning Strategies and Related Loss for Large Data

### 2015-02-18 Paper.Rnw
Spell check and minor grammar repair

Histograms are an example of plots which use binned variables. To extend histograms to a graph which can represent the joint distribution between two variables, it is a natural step to form a tesselated grid on a two dimensional Cartesian plane and use some other attribute (color, 3D rendering) to provide joint density information within each grid cell, known as a tile. A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) centered at the bin center, rather like a two-dimensional histogram viewed from above.

@@

\begin{figure}[hbtp]
  \centering
  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/Binning-DataAndBin.pdf}}
%   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
%   {./figure/RandomBinningDataPoints1}}
  \subfloat[Random Algorithm]{\includegraphics[keepaspectratio=TRUE, width=3in]
  {./figure/RandomBinningDataPoints2.pdf}}
  \caption{Points are assigned to bins centered at (5, 25) an (15,25). Points in the standard algorithm are contained within the defined bin region, where points from the random algorith may be assigned to bins from outside the defined bin regions.}
  \label{fig:BinningTutorial}
\end{figure}

@@

We could also consider a quantile based binning algorithm where bin widths are adjusted such that each bin contains an equal proportion of the observed values. This is a fundimentally different approach than the standard and random binning algorithms defined previously where the goal was to display approximate density. This binning will not be considered for binning the data directly, but instead for further binning the initial bin counts for the purposes of rendering colors on a discrete color palette.

Suppose that we have values $c_j$ for $j \in \{1,2,...J\}$ that we wish to divide into quantile based bins. The number of quantile based bins, $N_q ,= J$ , must be defined in order to assign values to bin centers. The quantile binning function $b^q_C(\cdot)$ assigns ordered values $c_{(j)}$ to bin centers $c^\ast_{(j)}$ as 
%
\begin{eqnarray}\label{quantbin}
c^\ast_{(j)} = b^q_C(c_{(j)}) = c_{(\eta)} \text{, where } \eta = \left\lfloor j/ \left\lceil J/N_c \right\rceil \right\rfloor \left\lceil J/N_c \right\rceil ,
\end{eqnarray}
%
\noindent and $\left\lceil \cdot \right\rceil$ is the ceiling function that rounds any real number up to the next highest integer.  To better understand this function we notice that $\left\lceil J/N_c \right\rceil$ represents the number of values to be assigned to each quantile bin, $n_q$, rounded up; thus the function assigns the $j^{th}$ value to the next lower ordered value that is a multiple of $n_q$.

@@

Therefore we distinguish two contributions to the total loss. 
Numerical loss results from the binning process as points are reduced to the numerical center point. The loss for an individual point, $L_i^\text{N}$, is the euclidean distance between the $i^{th}$ data point $(x_i,y_i)$ and the corresponding bin numerical center $(\bar{x}, \bar{y})$. The overall numerical loss is thus the sum of numerical losses for individual points ($L^\text{N} = \sum_{i=1}^n L_i^\text{N}$).
Visual loss is due to the distance between the numerical centers and the visual centers. The visual loss for each bin, $L_j^\text{V}$, is the euclidean distance between the visual center for the $j^{th}$ bin $(x_j^\ast, y_j^\ast)$ and the numerical center $(\bar{x}_j, \bar{y}_j)$. The overall visual loss is the sum of these bin distances ($L^\text{V} = \sum_{j=1}^{N_x} L_j^\text{V}$). 
This discrepancy is particularly clear as demonstrated in Figure~\ref{fig:stdBinTutorial2}, in which the visual center of the bin varies significantly from the numerical center of the data contained within the bin under the standard algorithm.  
This discrepancy between the visual center of the bin and the numerical center of the data within the bin occurs in both binning algorithms, and we can note that the loss functions napply to either approach to binning.

@@

Frequency loss is argueably the most important source of loss, as it is perceivable in the variability of the shading of bins which has a large effect on the utility of the graphical data summary. The standard and random algorithms are very similar when we compare frequency loss. The random algorithm has more variability in allocating observed values to bin centers, but the impact of this variability on the loss frequency information is much less noticable than the increase seen in numerical loss. %~\ref{fig:LossComparison}

Investigation of the loss sources we might be tempted to conclude that the standard binning algorithm is superior due to lower loss, however the random binning algorithm displays strong advantages when encountering problematic data structure issues. Random binning is especially advantagous when discrete data values are recorded at a consistant increments but we bin our data using a binwidth that is a non-integer multiple of the incremental widths. In this senario, we see that the non-synchronous data increments and bin widths will cause bins to contain unconsistant numbers of possible data value. Assuming that data values are spread uniformly over the discrete values we want a visualization that displays uniform frequency. Using the standard binning algorithm the misalignment of data increment and bin widths will lead to and artifical frequency discrepancy that manifests in a striped pattern in the visualization. The random binning algorithm is expected to display the intended uniform frequency pattern because the randomization of data values to bin centers mitigates the problematic bin width selection.

@@

For this data example we will revisit the baseball data used earlier in this paper.  The characteristics of loss described in the discussion above are exemplified by viewing the loss for the baseball data in Figure ~\ref{fig:LossComparison1}. We find that the visual and numerical loss are in fact higher while using the random binning algorithm, but that the two binning algorithms are almost indestiguishable in terms of frequency loss.

Another thing to note about Figure ~\ref{fig:LossComparison1} is that under the standard binning algorithm the numerical loss is a monotone increasing function of the bin width, but visual loss is not a strictly increasing function of bin width.  This is appearant in the order of colored lines based on the Strike Out binning size.  This is because the numerical loss function sums loss over all observations, which means the distance from the points to the numerical center of bins can only increase as binwidth increases. Whereas the visual loss function sums loss over bin, thus as the binwidth increases the number of bins decreases which leads to the potential that the overall visual loss may decrease even though the distance from numerical centers to visual centers may increase for individual bins.

@@

We may also use this example to highlight the advantages of random binning. The two binned scatterplots at the top of Figure~\ref{fig:binning-comparison} show bins of the same width, but clear differences between the binning results are evident. The standard binning algorithm has rounding artifacts that are clearly visible in the vertical strips. These vertical stripes are from the computational rounding issues encountered if we use the standard binning when we have the data values were recorded on an evenly spaces discrete scale (in our case, integers) and the binwidths are integer multiples of the data increments.  This reinforces the arguement that in practice the random algorithm can produce a graph that better represents the true shape of the data than the standard algorithm even if the numerical and visual loss is higher.

>>>>

Histograms are an example of plots which use binned variables. To extend histograms to a graph which can represent the joint distribution between two variables, it is a natural step to form a tessellated grid on a two dimensional Cartesian plane and use some other attribute (color, 3D rendering) to provide joint density information within each grid cell, known as a tile. A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) centered at the bin center, rather like a two-dimensional histogram viewed from above.

@@

\begin{figure}[hbtp]
  \centering
  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/Binning-DataAndBin.pdf}}
%   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
%   {./figure/RandomBinningDataPoints1}}
  \subfloat[Random Algorithm]{\includegraphics[keepaspectratio=TRUE, width=3in]
  {./figure/RandomBinningDataPoints2.pdf}}
  \caption{Points are assigned to bins centered at (5, 25) an (15,25). Points in the standard algorithm are contained within the defined bin region, where points from the random algorithm may be assigned to bins from outside the defined bin regions.}
  \label{fig:BinningTutorial}
\end{figure}

@@

We could also consider a quantile based binning algorithm where bin widths are adjusted such that each bin contains an equal proportion of the observed values. This is a fundamentally different approach than the standard and random binning algorithms defined previously where the goal was to display approximate density. This binning will not be considered for binning the data directly, but instead for further binning the initial bin counts for the purposes of rendering colors on a discrete color palette.

Suppose that we have values $c_j$ for $j \in \{1,2,...J\}$ that we wish to divide into quantile based bins. The number of quantile based bins, $N_q ,= J$ , must be defined in order to assign values to bin centers. The quantile binning function $b^q_C(\cdot)$ assigns ordered values $c_{(j)}$ to bin centers $c^\ast_{(j)}$ as 
%
\begin{eqnarray}\label{quantbin}
c^\ast_{(j)} = b^q_C(c_{(j)}) = c_{(\eta)} \text{, where } \eta = \left\lfloor j/ \left\lceil J/N_c \right\rceil \right\rfloor \left\lceil J/N_c \right\rceil ,
\end{eqnarray}
%
\noindent and $\left\lceil \cdot \right\rceil$ is the ceiling function that rounds any real number up to the next highest integer.  To better understand this function we notice that $\left\lceil J/N_c \right\rceil$ represents the number of values to be assigned to each quantile bin, $n_q$, rounded up; thus $\eta$ is the index for the ordered value where the next lower quantile based bin will be centered.

@@

Therefore we distinguish two contributions to the total loss. 
Numerical loss results from the binning process as points are reduced to the numerical center point. The loss for an individual point, $L_i^\text{N}$, is the euclidean distance between the $i^{th}$ data point $(x_i,y_i)$ and the corresponding bin numerical center $(\bar{x}, \bar{y})$. The overall numerical loss is thus the sum of numerical losses for individual points ($L^\text{N} = \sum_{i=1}^n L_i^\text{N}$).
Visual loss is due to the distance between the numerical centers and the visual centers. The visual loss for each bin, $L_j^\text{V}$, is the euclidean distance between the visual center for the $j^{th}$ bin $(x_j^\ast, y_j^\ast)$ and the numerical center $(\bar{x}_j, \bar{y}_j)$. The overall visual loss is the sum of these bin distances ($L^\text{V} = \sum_{j=1}^{N_x} L_j^\text{V}$). 
This discrepancy is particularly clear as demonstrated in Figure~\ref{fig:stdBinTutorial2}, in which the visual center of the bin varies significantly from the numerical center of the data contained within the bin under the standard algorithm.  
This discrepancy between the visual center of the bin and the numerical center of the data within the bin occurs in both binning algorithms, and we can note that the loss functions apply to either approach to binning.

@@

Frequency loss is arguably the most important source of loss, as it is perceivable in the variability of the shading of bins which has a large effect on the utility of the graphical data summary. The standard and random algorithms are very similar when we compare frequency loss. The random algorithm has more variability in allocating observed values to bin centers, but the impact of this variability on the loss frequency information is much less noticeable than the increase seen in numerical loss. %~\ref{fig:LossComparison}

Investigation of the loss sources we might be tempted to conclude that the standard binning algorithm is superior due to lower loss, however the random binning algorithm displays strong advantages when encountering problematic data structure issues. Random binning is especially advantageous when discrete data values are recorded at a consistent increments but we bin our data using a binwidth that is a non-integer multiple of the incremental widths. In this scenario, we see that the non-synchronous data increments and bin widths will cause bins to contain inconstant numbers of possible data value. Assuming that data values are spread uniformly over the discrete values we want a visualization that displays uniform frequency. Using the standard binning algorithm the misalignment of data increment and bin widths will lead to and artificial frequency discrepancy that manifests in a striped pattern in the visualization. The random binning algorithm is expected to display the intended uniform frequency pattern because the randomization of data values to bin centers mitigates the problematic bin width selection.

@@

For this data example we will revisit the baseball data used earlier in this paper.  The characteristics of loss described in the discussion above are exemplified by viewing the loss for the baseball data in Figure ~\ref{fig:LossComparison1}. We find that the visual and numerical loss are in fact higher while using the random binning algorithm, but that the two binning algorithms are almost indistinguishable in terms of frequency loss.

Another thing to note about Figure ~\ref{fig:LossComparison1} is that under the standard binning algorithm the numerical loss is a monotone increasing function of the bin width, but visual loss is not a strictly increasing function of bin width.  This is apparent in the order of colored lines based on the Strike Out binning size.  This is because the numerical loss function sums loss over all observations, which means the distance from the points to the numerical center of bins can only increase as binwidth increases. Whereas the visual loss function sums loss over bin, thus as the binwidth increases the number of bins decreases which leads to the potential that the overall visual loss may decrease even though the distance from numerical centers to visual centers may increase for individual bins.

@@

We may also use this example to highlight the advantages of random binning. The two binned scatterplots at the top of Figure~\ref{fig:binning-comparison} show bins of the same width, but clear differences between the binning results are evident. The standard binning algorithm has rounding artifacts that are clearly visible in the vertical strips. These vertical stripes are from the computational rounding issues encountered if we use the standard binning when we have the data values were recorded on an evenly spaces discrete scale (in our case, integers) and the binwidths are integer multiples of the data increments.  This reinforces the argument that in practice the random algorithm can produce a graph that better represents the true shape of the data than the standard algorithm even if the numerical and visual loss is higher.
